---
title: "Boreal forest (subarctic climate) burned area 2002–2021"
author: "Manuel Popp"
date: "2024-12-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, error = FALSE
  )

```

## Introduction

In this document, we will explore burned area in subarctic climate (Köppen-Geiger classes *Dfc, Dwc, Dsc, Dfd, Dwd* and *Dsd*) between 2002 and 2021 in the context of shifting boundaries of this climate zone.

### Data sets
- MODIS burned area
- Modis landcover product MCD12Q1
- CHELSA Köppen-Geiger classification for two periods:
-- 1951–1980
-- 1981–2010

```{r Load packages, include=FALSE, message=FALSE, warnings=FALSE}
rm(list = ls())
import <- function(...) {
  #' Import R packages. Install them if necessary.
  #' 
  #' @param ... any argument that can be passed to install.packages.
  #' @details The function installs only packages that are missing. Packages
  #' are loaded.
  #' @examples
  #' # Load packages
  #' import("dplyr", "MASS", "terra", dependencies = TRUE)
  #' 
  #' @seealso \code{\link[base]{install.packages}}
  #' @export
  args <- list(...)
  packages = args[names(args) == ""]
  kwargs = args[names(args) != ""]
  
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      do.call(install.packages, c(list(package), kwargs))
    }
    require(package, character.only = TRUE)
  }
}

import(
  "terra", "dplyr", "tidyterra", "progress", "car", "MASS", "glmtoolbox",
  "performance", "ggplot2", "RSpectra", "spaMM", "mgcv",
  dependencies = TRUE
)

```


```{r Settings, echo=FALSE, message=FALSE, warnings=FALSE}
recalculate <- FALSE
options(terra.progress = 0)

if (Sys.info()["sysname"] == "Windows") {
  dir_main <- "C:/Users/poppman/switchdrive/PhD/prj/bff"
  sub_clim <- "chelsa_kg"
} else {
  dir_main <- "/lud11/poppman/data/bff"
  sub_clim <- file.path("lud11", "chelsa_kg")
}

dir_dat <- file.path(dir_main, "dat")
dir_lud <- file.path(dir_dat, "lud11")
dir_clim <- file.path(dir_dat, sub_clim)
dir_bor <- file.path(dir_dat, "boreal")

f_kg1951 <- file.path(dir_clim, "CHELSA_kg0_1951-1980_V.2.1.tif")
f_kg1981 <- file.path(dir_clim, "CHELSA_kg0_1981-2010_V.2.1.tif")
f_kg1951m <- file.path(dir_bor, "CHELSA_kg0_1951-1980_masked.tif")
f_kg1981m <- file.path(dir_bor, "CHELSA_kg0_1981-2010_masked.tif")
f_bor1951 <- file.path(dir_bor, "boreal_kg0_1951-1980.tif")
f_bor1981 <- file.path(dir_bor, "boreal_kg0_1981-2010.tif")
f_boreal_poly <- file.path(dir_bor, "boreal_kg0.gpkg")
f_boreal_samples <- file.path(dir_bor, "boreal_samples.gpkg")
f_needleforest <- file.path(dir_bor, "evergr_needleleaf_mask.tif")
f_mask <- file.path(dir_dat, "mask", "combined_mask.tif")
f_fire <- file.path(dir_dat, "fire_freq", "Fire_freq2002-2021.tif")
dir_chelsa <- file.path(dir_dat, "chelsa")

subarc_classnums <- c(20, 21, 24, 25, 28, 29)
n_samples <- 5e3

wsl_cols <- c(
  rgb(0, 102, 102, maxColorValue = 255),
  "skyblue3"
)

```

Firt, we load the fire raster as a template for resampling all other rasters.

```{r, message=FALSE, warnings=FALSE}
fire <- terra::rast(f_fire)

```

Second, we create a mask of water bodies and human modification to exclude pixels that are covered by water or infrastructure, or are extensively managed.
```{r, message=FALSE, warnings=FALSE}
 if (!file.exists(f_mask) | recalculate) {
  f_gHM <- file.path(dir_lud, "GlobalHumanModification", "gHM_4326.tif")
  f_lcv <- file.path(dir_lud, "modis", "landcover", "MCD12Q1_water_4326.tif")
  
  human <- terra::rast(f_gHM) %>%
    terra::resample(fire, method = "near")
  water <- terra::rast(f_lcv) %>%
    terra::resample(fire, method = "near") %>%
    terra::classify(rcl = matrix(c(1, NA), ncol = 2))
  
  combined_mask <- terra::mask(
    human, water, maskvalue = NA, updatevalue = NA
    ) %>%
    terra::writeRaster(
      filename = f_mask,
      datatype = "INT1U",
      gdal = c("COMPRESS=DEFLATE", "TFW=YES"),
      overwrite = TRUE,
      NAflag = 0
    )
  
  rm(human)
  rm(water)
  gc()
} else {
  combined_mask <- terra::rast(f_mask)
}

```

Now, we resample the Köppen-Geiger climate rasters to match the extent and resolution of the MODIS fire product. Moreover, we identify regions of subarctic climate for both climate periods and we derive regions of stable climate, as well as regions classified as "subarctic climate" during the first, but not during the second period. In the following, regions of stable climate will be labelled "stable", while we use the term "loss" to describe regions that experienced a shift from a subarctic to any non-subarctic Köppen-Geiger climate class.

```{r, message=FALSE, warnings=FALSE}
if (!file.exists(f_kg1951m) | recalculate) {
  kg1951 <- terra::rast(f_kg1951) %>%
    terra::resample(fire, method = "near") %>%
    terra::mask(combined_mask) %>%
    terra::writeRaster(
      filename = f_kg1951m,
      datatype = "INT2U",
      gdal = c("COMPRESS=DEFLATE", "TFW=YES"),
      overwrite = TRUE,
      NAflag = 0
    )
} else {
  kg1951 <- terra::rast(f_kg1951m)
}

if (!file.exists(f_kg1981m) | recalculate) {
  kg1981 <- terra::rast(f_kg1981) %>%
    terra::resample(fire, method = "near") %>%
    terra::mask(combined_mask) %>%
    terra::writeRaster(
      filename = f_kg1981m,
      datatype = "INT2U",
      gdal = c("COMPRESS=DEFLATE", "TFW=YES"),
      overwrite = TRUE,
      NAflag = 0
    )
} else {
  kg1981 <- terra::rast(f_kg1981m)
}

if (!file.exists(f_bor1951) | !file.exists(f_bor1981) | recalculate) {
  classes <- unique(
    c(terra::unique(kg1951), terra::unique(kg1981))
  )[[1]]
  
  reclass <- cbind(classes, rep(NA, length(classes)))
  reclass[which(reclass[, 1] %in% subarc_classnums), 2] <- seq(
    1, length(subarc_classnums)
  )
  
  bor1951 <- kg1951 %>%
    terra::classify(rcl = reclass) %>%
    terra::writeRaster(
      filename = f_bor1951,
      datatype = "INT1U",
      gdal = c("COMPRESS=DEFLATE", "TFW=YES"),
      overwrite = TRUE,
      NAflag = 0
    )
  
  bor1981 <- kg1981 %>%
    terra::classify(rcl = reclass) %>%
    terra::writeRaster(
      filename = f_bor1981,
      datatype = "INT1U",
      gdal = c("COMPRESS=DEFLATE", "TFW=YES"),
      overwrite = TRUE,
      NAflag = 0
    )
} else {
  bor1951 <- terra::rast(f_bor1951)
  bor1981 <- terra::rast(f_bor1981)
}

boreal_union <- terra::not.na(terra::cover(bor1951, bor1981), falseNA = TRUE)
boreal_union_classified <- (
  terra::not.na(bor1951, falseNA = FALSE) +
    (terra::not.na(bor1981, falseNA = FALSE) * 2)
  ) %>%
  terra::classify(rcl = matrix(c(0, 1, 2, 3, NA, 1, NA, 2), ncol = 2))# Only loss and stable

```

We use the combined area of subarctic climate to mask the fire product:
```{r, message=FALSE, warnings=FALSE}
fire <- terra::rast(f_fire) %>%
  terra::mask(boreal_union, maskvalue = NA, updatevalue = NA)

```

Finally, we conduct a stratified sampling to select the same number of pixels from both, loss and stable climate area. To exclude non-forest pixels, we first mask the subarctic climate with a forest mask derived from the MODIS landcover product PFT layer (evergreen and deciduous needle leaf forest).
```{r, message=FALSE, warnings=FALSE, error=TRUE}
forest_mask <- terra::rast(f_needleforest) %>%
  terra::resample(boreal_union_classified, method = "near")

boreal_union_classified <- terra::mask(
  boreal_union_classified, mask = forest_mask,
  maskvalue = NA, updatevalue = NA
)

tryCatch(
  {# This might fail due to large data size.
    samples <<- terra::spatSample(
      boreal_union_classified, size = n_samples, method = "stratified",
      replace = FALSE, na.rm = TRUE, as.points = TRUE, values = FALSE,
      exhaustive = TRUE
    )
  },
  error = function(cond) {
    message("Failed to sample pixels:")
    message(conditionMessage(cond))
    
    sample_list <- list()
    for (class in terra::unique(boreal_union_classified)[, 1]) {
      idx <- which(terra::values(boreal_union_classified, mat = FALSE) == class)
      ilocs <- sample(idx, size = n_samples)
      sample_list[[class]] <- terra::vect(
        terra::xyFromCell(boreal_union_classified, cell = ilocs),
        crs = "epsg:4326"
      )
    }
    
    samples <<- terra::vect(do.call(c, sample_list))
  },
  finally = {
    values <- terra::extract(boreal_union_classified, samples)
    cat(
      "\nSample length:", length(samples), "\nof which",
      length(which(values[, 2] == 1)), "are class 'loss' and",
      length(which(values[, 2] == 2)), "are class 'stable'.\n"
      )
  }
)

sample <- c(fire, bor1951, bor1981) %>%
  terra::extract(y = samples) %>%
  purrr::set_names(c("ID", "fire", "kg1951", "kg1981")) %>%
  dplyr::mutate(
    type = ifelse(
      !is.na(kg1951) & !is.na(kg1981), "stable",
      ifelse(
        is.na(kg1951),
        "gain",
        "loss"
        )
      )
    )

```

Now, we add climate information from CHELSA.
```{r, message=FALSE, warnings=FALSE}
files_chelsa_vars <- list.files(
  dir_chelsa, pattern = ".tif$", full.names = TRUE
)

chelsa_names <- sub(".tif", "", basename(files_chelsa_vars))

dir_chelsa_resampled <- paste0(dir_chelsa, "_resampled")

dir.create(dir_chelsa_resampled, showWarnings = FALSE)

for (file in files_chelsa_vars) {
  file_resampled <- file.path(dir_chelsa_resampled, basename(file))
  
  if (!file.exists(file_resampled) | recalculate) {
    terra::rast(file) %>%
      terra::resample(fire, method = "near") %>%
      terra::mask(boreal_union, maskvalue = NA, updatevalue = NA) %>%
      terra::writeRaster(filename = file_resampled, overwrite = TRUE)
  }
  
  gc()
}

# Use resampled CHELSA variables
files_chelsa_resampled <- list.files(
  dir_chelsa_resampled, pattern = ".tif$", full.names = TRUE
)
chelsa_names <- sub(".tif", "", basename(files_chelsa_resampled))

chelsa <- do.call(c, lapply(files_chelsa_resampled, terra::rast))
names(chelsa) <- chelsa_names

data <- c(fire, bor1951, bor1981, chelsa) %>%
  terra::extract(y = samples) %>%
  purrr::set_names(c("ID", "fire", "kg1951", "kg1981", chelsa_names)) %>%
  dplyr::mutate(
    type = ifelse(
      !is.na(kg1951) & !is.na(kg1981), "stable",
      ifelse(
        is.na(kg1951),
        "gain",
        "loss"
      )
    )
  )

data <- data[which(data$type %in% c("loss", "stable")),]
data$type <- factor(data$type)

```

Fit a generalised linear model (GLM).
```{r}
glm_formula <- as.formula(
  paste(
    "fire ~ type", " * (",
    paste(chelsa_names, collapse = " + "), ")",
    sep = ""
    )
  )

glm_null_formula <- as.formula(
  paste("fire ~ ", paste(chelsa_names, collapse = " + "), sep = "")
)

mod_interaction <- glm(
  formula = glm_formula, data = data, family = poisson(link = "log")
  )

mod_null <- glm(
  formula = glm_null_formula, data = data, family = poisson(link = "log")
  )

# Check for overdispersion
overdispersion <- performance::check_overdispersion(mod_interaction)
if (
  overdispersion$p_value < 0.1
  ) {
  mod_interaction <- MASS::glm.nb(formula = glm_formula, data = data)
  mod_null <- MASS::glm.nb(formula = glm_null_formula, data = data)
}

summary(mod_interaction)

# Compare effects
aov_full <- anova(mod_interaction, mod_null)
print(aov_full)

#------------------------------------------------------------------------------|
# Optimise model
step_out <- glmtoolbox::stepCriterion(
  model = mod_interaction,
  test = "lr",
  criterion = "p-value",
  direction = "backward"
)

#------------------------------------------------------------------------------|
# Compute GLM again with relevant predictors
glm_final_formula <- as.formula(paste0("fire", step_out$final))
mod_glm <- glm(
  formula = glm_final_formula, data = data, family = poisson(link = "log")
)

overdispersion <- performance::check_overdispersion(mod_interaction)
if (
  overdispersion$p_value < 0.1
) {
  mod_glm <- MASS::glm.nb(
    formula = glm_final_formula, data = data, family = poisson(link = "log")
  )
  mod_null <- MASS::glm.nb(formula = glm_null_formula, data = data)
}

# Compare effects
options(contrasts = c("contr.sum", "contr.poly"))
car::Anova(mod_glm, type = "III")
aov_glm <- anova(mod_glm, mod_null)
print(aov_glm)

```

We observe that "type", i.e., whether a pixel experienced a shift into another Köppen-Geiger climate class or not, is borderline significant. However, interactions of this factor are mostly highly significant.

One of these interactions is that with annual surface temperature. We plot the relationship between temperature and number of burned pixels within both, "loss" and "stable" climate area, as well as the fitted GLM:
```{r, echo=FALSE, message = FALSE, warning = FALSE}
ggplot2::ggplot(data, aes(x = tas_mean, y = fire)) +
  ggplot2::geom_point(alpha = 0.6, colour = wsl_cols[1]) +
  ggplot2::geom_smooth(
    method = "glm.nb", se = TRUE, colour = wsl_cols[2]
    ) +
  ggplot2::facet_wrap(~ type) +
  ggplot2::labs(
    title = "Pixel-wise fire count across mean annual temperture by region",
    caption = paste(
      "Figure 1.",
      "Counted number of burned pixels between 2002 and 2021.",
      "Blue lines show the GLM fit, grey background",
      "indicates the confidence intervals."
      )
    ) +
  ggplot2::xlab("Mean air temperature during the time period") +
  ggplot2::ylab("Number of pixels burned during time period") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.caption = ggplot2::element_text(hjust = 0)
  )

```

Since all samples were taken from a spatial raster, they are not independent samples! A fire does not automatically ignite once fuel accumulated and the weather is suitable. Even under good fire conditions, it might be highly unlikely, since it requires some spark. However, once an adjacent pixel is burning, the fire risk is skyrocketing. We try to account for that by using a Matérn term:

```{r, message = FALSE, warning = FALSE}
# Control for spatial proximity of pixels
row_layer <- terra::init(chelsa, fun = "row")
col_layer <- terra::init(chelsa, fun = "col")
coords <- c(row_layer, col_layer) %>%
  terra::extract(y = samples) %>%
  as.data.frame()

names(coords) <- c("id", "y", "x")
df <- cbind(data, coords)

#-------------------------------------------------------------------------------
# Run with a subset for testing and fast computation of a preliminary result
save(df, file = file.path(dir_bor, "df.Rdata"))

subset_size <- 1500
idx <- c(
  sample(which(df$type == levels(df$type)[1]), subset_size),
  sample(which(df$type == levels(df$type)[2]), subset_size)
)

df <- df[idx,]
#-------------------------------------------------------------------------------

f_mod_matern <- file.path(dir_bor, "mod_matern.Rsave")

if (!file.exists(f_mod_matern) | recalculate) {
  num_cores <- parallel::detectCores(logical = FALSE)
  
  mod_matern_full <- spaMM::fitme(
    fire ~ type + cmi_dry_rate + cmi_mean + pr_mean + swb_mean + 
      tas_max + tas_mean + tas_min + type:cmi_dry_rate + type:tas_max + 
      type:tas_mean + type:tas_min + Matern(1 | x + y),
    data = df, family = spaMM::negbin(),
    control.HLfit = list(
      algebra = "decorr",# "decorr" has been determined in first run.
      NbThreads = ifelse(num_cores > 3, num_cores - 2L, 1L)
    )
  )
  
  mod_matern_null <- spaMM::fitme(
    fire ~ cmi_dry_rate + cmi_mean + pr_mean + swb_mean + tas_max + tas_mean +
      tas_min + Matern(1 | x + y),
    data = df, family = spaMM::negbin(),
    control.HLfit = list(
      algebra = "decorr",# "decorr" has been determined in first run.
      NbThreads = ifelse(num_cores > 3, num_cores - 2L, 1L)
    )
  )
  
  save(mod_matern_full, mod_matern_null, file = f_mod_matern)
 } else {
  load(f_mod_matern)
}

lrt <- spaMM::LRT(mod_matern_full, mod_matern_null)
print(lrt)

```

Now, we can use the fit model to estimate fire risk and print the estimates in a coplot with important predictors on the x-axis, the type (loss or stable climate) as facets, and (predicted) fire frequency on the y-axis.

First, we look at the mean temperature of the hottest year:
```{r, echo=FALSE, message = FALSE, warning = FALSE}
#...............................................................................
# Calculate mean values for additional predictors
mean_values <- data.frame(lapply(df, mean, na.rm = TRUE))
x_values <- expand.grid(
  tas_max = seq(min(df$tas_max), max(df$tas_max), length.out = 100),
  type = unique(df$type)
)

new_data <- cbind(
  x_values, mean_values[, which(!names(mean_values) %in% names(x_values))]
  )

# Predict from the model
new_data$predicted_fire <- as.numeric(
  predict(mod_matern_full, newdata = new_data, type = "response")
)

gg_1 <- ggplot2::ggplot(data, aes(x = tas_max / 10, y = fire)) +
  ggplot2::geom_point(alpha = 0.3, colour = wsl_cols[1]) +
  # ggplot2::geom_smooth(
  #   method = "glm.nb", se = TRUE, colour = wsl_cols[2]
  #   ) +
  ggplot2::geom_smooth(
    data = new_data, aes(y = predicted_fire), colour = wsl_cols[2]
    ) +
  ggplot2::facet_wrap(~ type) +
  ggplot2::labs(
    title = "Coplot of fire frequency and temperature for stable and lost climate area",
    caption = paste(
      "Figure 1.",
      "Number of fire detections per pixel between 2002 and 2021.",
      "Blue lines display GLM predictions (negative binomial)."
      )
    ) +
  ggplot2::xlab("Mean air temperature (K) of the hottest year") +
  ggplot2::ylab("Detected fires per pixel") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.caption = ggplot2::element_text(hjust = 0)
  )

plot(gg_1)

```

Let's look at the best predictor (overall average temperature):
```{r, echo=FALSE, message = FALSE, warning = FALSE}
mean_values <- data.frame(lapply(df, mean, na.rm = TRUE))
x_values <- expand.grid(
  tas_mean = seq(min(df$tas_mean), max(df$tas_mean), length.out = 100),
  type = unique(df$type)
)

new_data <- cbind(
  x_values, mean_values[, which(!names(mean_values) %in% names(x_values))]
)

# Predict from the model
new_data$predicted_fire <- as.numeric(
  predict(mod_matern_full, newdata = new_data, type = "response")
)

gg_2 <- ggplot2::ggplot(
  data, aes(x = tas_mean / 10, y = fire)
  ) +
  ggplot2::geom_point(alpha = 0.3, colour = wsl_cols[1]) +
  # ggplot2::geom_smooth(
  #   method = "glm.nb", se = TRUE, colour = wsl_cols[2]
  # ) +
  ggplot2::geom_smooth(
    data = new_data, aes(y = predicted_fire), colour = wsl_cols[2]
  ) +
  ggplot2::facet_wrap(~ type) +
  ggplot2::labs(
    title = "Coplot of fire frequency and temperature for stable and lost climate area",
    caption = paste(
      "Figure 2.",
      "Number of fire detections per pixel between 2002 and 2021.",
      "Blue lines display GLM predictions (negative binomial)."
    )
  ) +
  ggplot2::xlab("Mean air temperature (K) during the measurement period") +
  ggplot2::ylab("Detected fires per pixel") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.caption = ggplot2::element_text(hjust = 0)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 5))

plot(gg_2)

```

Fit a GAM with spatial term...
```{r, warning = FALSE}
df <- cbind(data, coords)
mod_gam <- mgcv::gam(
  fire ~ type + s(cmi_dry_rate) + s(cmi_mean) + s(pr_mean) + s(swb_mean) + 
    s(tas_max) + s(tas_mean) + s(tas_min) + s(x, y, bs = "gp"),
  family = poisson(link = "log"),
  data = df,
  method = "REML"
)

summary(mod_gam)
plot(mod_gam, pages = 1)
gam.check(mod_gam)

```